---
title: "Exploratory Data Analysis"
author: "Rachel Roggenkemper, Jacob Perez, Brendan Callender"
format: html
editor: visual
code-fold: true
---

# Setup & Data Cleaning

```{r}
#| label: load-packages

library(tidyverse)
```

```{r}
#| label: read-data
#| message: false

cvd <- read_csv(here::here("data", "cvd_data.csv"))
diab <- read_csv(here::here("data", "diabetes_full.csv"))
es_diab <- read_csv(here::here("data", "early_stage_diabetes.csv"))
```

```{r}
#| label: clean-response-vars

cvd <- cvd %>%
  mutate(target = if_else(target == 0, "no heart disease", "heart disease"),
         target = factor(target, levels = c("heart disease", "no heart disease")))

diab <- diab %>%
  mutate(
    target = case_when(
      Diabetes_012 == 0 ~ "no diabetes",
      Diabetes_012 == 1 ~ "diabetes", # prediabetes
      Diabetes_012 == 2 ~ "diabetes"
  ),
    target = factor(target, levels = c("diabetes", "no diabetes"))
  ) %>%
  select(-Diabetes_012)

es_diab <- es_diab %>%
  mutate(target = if_else(class == "Negative", "no diabetes", "diabetes"),
         target = factor(target, levels = c("diabetes", "no diabetes"))) %>%
  select(-class)
```

# Distribution of Response Variables

## Cardiovascular Disease Dataset

```{r}
N <- nrow(cvd)
cvd %>%
  select(target) %>%
  group_by(target) %>%
  summarize(n = n()) %>%
  mutate(prop = n / N)
```

## Early Stage Diabetes

```{r}
N <- nrow(es_diab)
es_diab %>%
  select(target) %>%
  group_by(target) %>%
  summarize(n = n()) %>%
  mutate(prop = n / N)
```

## Diabetes CDC

```{r}
N <- nrow(diab)
diab %>%
  select(target) %>%
  group_by(target) %>%
  summarize(n = n()) %>%
  mutate(prop = n / N)
```

# Exploratory Data Analysis

Cool idea: Attempt to make two models for men and women separately and compare to 1 single model. Men and women have different indicators for certain diseases so doing this could help account for that.

## Cardiovascular Disease

```{r}
cvd %>%
  ggplot(aes(x = age, fill = target)) +
  geom_histogram(bins = 10) +
  facet_wrap(~target) +
  theme(legend.position = "none")
```

```{r}
cvd %>%
  ggplot(aes(x = maxheartrate, y = restingBP, color = target)) +
  geom_point()
```
